Density-dependent selection and the limits of relative fitness

### 2018-07-11 manuscript.tex
latest

There are a variety of different measures of fitness, such as expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is widely used in evolutionary genetics, where the focus is on relative genotypic frequencies \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, but it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp. 178]{charlesworth_1994}). While such a connection is clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the widespread use of relative fitness in evolutionary genetics, it is important to understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

@@

Each iteration, adults produce propagules which disperse at random, independently of distance from their parents, and independently of each other (undirected dispersal). We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$. The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fertility and dispersal ability \citep{levins_71,tilman_94}. Random dispersal is then modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density in unoccupied territories. The total propagule density will be denoted $L=\sum_i l_i$.

@@

Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model, because greater $b$ not only means more propagules contesting territories, but also more territories being contested. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection). During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).

>>>>

There are a variety of different measures of fitness, such as expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is widely used in evolutionary genetics, where the focus is on relative genotypic frequencies \cite[pp. 468]{barton_2007}. The justification of any measure of fitness ultimately derives from how it is connected to the processes of birth and death which drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp. 178]{charlesworth_1994}). While such a connection is clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the widespread use of relative fitness in evolutionary genetics, it is important to understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

@@

Each iteration, adults produce propagules which disperse at random, independently of distance from their parents, and independently of each other (undirected dispersal). We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$ (the factor $U/T$ represents the loss of those propagules landing on occupied territories). Random dispersal is modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density of type $i$ per unoccupied territory. The total propagule density per unoccupied territory will be denoted $L=\sum_i l_i$.

@@

Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model. Greater $b$ means not only that more propagules contest the available territories, but also that a greater fraction of unoccupied territories receive propagules. Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection). During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details). Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).
